Stochastic Wind Power Forecasting

### 2020-03-22 LaTeX_code/paper/paper_1line.tex
Section 2 almost done

\begin{itemize}
\item $a(\cdot; p_t, \dot{p}_t, \bm{\theta}): [0,1] \to \mathbb{R} $  denotes a drift function,
\item $b(\cdot; p_t, \dot{p}_t, \bm{\theta}): [0,1] \to \mathbb{R}^+ $  a  diffusion function,
\item $\bm{\theta}$ is a vector of unknown parameters,
\item $(p_t)_{t \in [0,T]}$ is a time-dependent deterministic function $[0,1]$-valued and $ (\dot{p}_t)_{t \in [0,T]}$ is its time derivative,
\item $\{W_t, \, t \in [0,T] \}$ is a standard real-valued Wiener process.
\end{itemize}

@@

Let $\bm{\theta} = (\theta_0,\alpha)$, and choose a state-dependent diffusion term that avoids the process exiting from the range $[0,1]$ as follows:
  \begin{equation}
    b (X_t; \bm{\theta} )= \sqrt{2 \theta_0 \alpha X_t (1 - X_t)}
  \end{equation}
  where $\alpha >0$ is a unknown parameter that controls the path variability. This diffusion term belongs to the Pearson diffusion family and, in particular, it defines a Jacobi type diffusion. It is useful to recall that (\cite[440]{foso}) a Pearson diffusion is a stationary solution to a stochastic differential equation of the form
 \begin{equation}
    dX_t = - \theta (X_t - \mu) dt + \sqrt{2 \theta (a X_t^2 + b X_t + c)} dW_t
  \end{equation}
where $\theta>0$, and $a$, $b$ and $c$ are parameters such that the square root is well defined when $X_t$ is in the state space. These parameters, together with $\mu$, the mean of the invariant distribution, determine the state space of the diffusion as well as the shape of the invariant distribution.

An exhaustive classification of the (stationary) Pearson diffusions is presented in (\cite[440-443]{foso}) where, in particular, it is discussed the case $a < 0$ and $b(x; \bm{\theta}) = \sqrt{2 \theta a x (x-1)}$, where the invariant distribution is a Beta distribution with parameters $\left( \frac{\mu}{-a}, \frac{1 - \mu}{-a} \right),$ that leads to the well-known Jacobi diffusions, so-called because the eigenfunctions of the infinitesimal generator of these processes are the Jacobi polynomials (see, for example, \cite[2860-2861]{leph}).

It is worth mentioning that Jacobi diffusions have been successfully applied in several disciplines, among them econometrics (see  and references therein) and neuroscience (\cite{dotala}).

However, a distinctive feature in our proposed model 
\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dX_t \!\!\!&= (\dot{p}_t  - \theta_t (X_t - p_t) ) dt +\sqrt{2 \theta_0 \alpha X_t (1 - X_t)} dW_t\,, \quad t \in [0,T]  \\
   X_0  \!\!\!&=  x_0\in [0,1] \,,
 \end{array}\right.  \label{ourmodel}
\end{equation}

is that the drift term contains the time-varying parameter $\theta_t$, rendering the solution $X$ to (\ref{ourmodel}) a non-stationary and time-inhomogeneous process. To ensure that the process $X$ be the unique strong solution to (\ref{ourmodel}) for all $t \in [0,T]$ with state space $[0,1]$ a.s., the mean-reversion time-varying parameter must satisfy the following condition:
%has to be selected according to the following rule. Observe that the time derivative term $\dot{p}_t $ is not controlled to maintain that $X_t$ stays a.s. inside the range $[0,1]$. In other words, the zero drift line defined by $a(x; p_t,\bm{\theta}) =0$, which is an attractor, must be contained inside the range $[0,1]$. Thus, we must have that
%\begin{equation}
%\frac{- |\dot{p_t}|}{p_t} \leq \theta_t \leq \frac{|\dot{p}_t|}{1- p_t}
%\end{equation}
%which is satisfied  by choosing a time-dependent  $\theta_t$ as follows,
%\begin{equation}
%\theta_t = \max \left( \theta_0 \ , \ \frac{|\dot{p}_t|}{\min (p_t, 1-p_t)}  \right ),  \quad \theta_0 >0\,. \label{theta_t}
%\end{equation}
\begin{equation}\label{Assumption:3}
\theta_t\geq \max\left(\frac{\alpha\theta_0+\dot p_t}{1-p_t},\frac{\alpha\theta_0-\dot p_t}{p_t}\right)\tag{C}. 
\end{equation}
The proof of this theoretical statement is presented in Section \ref{Appendix}.

@@

%\subsubsection*{Change of Variables:}
\subsection{A model specification for the forecast error}
 After applying the simple change of variables $$V_t = X_t - p_t \,,$$ we may introduce the following model for the forecast error of the normalized wind power production:

\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dV_t \!\!\!&=  - \theta_t V_t  \, dt + \sqrt{2 \theta_t \alpha (V_t +p_t ) (1-V_t-p_t)} \, dW_t\,, \quad t \in [0,T]  \\
   V_0  \!\!\!&=  v_0\in [-1,1] \,.
 \end{array}\right.  \label{VtSDE}
\end{equation}

@@

We consider the following Lamperti transformation  {\color{red} (Remind that in continuous-time models with different diffusion term are singular and the likelihood cannot be written down. Besides, it can be unstable to work with low frequency. If the frequency of the data is too low, the Lamperti transform helps to detect this problem (in principle, it is not clear how to determine the right frequency for estimation. )}
\begin{equation}
Z_t = h(V_t) = \frac{1}{\sqrt{2 \theta_t \alpha}} \int_{\xi}^{x} \frac{1}{\sqrt{(u + p_t)(1 - u - p_t)}} \,du \Bigg{\vert}_{x = V_t} \,,
\end{equation}
where $\xi$ is an arbitrary point of the state space of the process $V$. The choice of $\xi = \frac{1}{2} - p_t$, where $p_t$ is a known input, leads to the process 
\begin{equation}
Z_t = h(V_t) = \frac{1}{\sqrt{2 \theta_t \alpha}} \arcsin (2(V_t + p_t) -1)\,,
\end{equation}
that, after applying It\^{o}'s formula on $h(V_t)$, gives the state-independent diffusion SDE
\begin{equation}
dZ_t = \left[  \frac{- \theta_t V_t}{ \sqrt{2 \theta_t \alpha (V_t + p_t)(1 - V_t - p_t)}} - \frac{1}{4} \frac{\sqrt{2 \theta_t \alpha} \left( 1 - 2 (V_t + p_t)\right)}{\sqrt{(V_t + p_t)(1 - V_t - p_t)}}  \right] \,dt + dW_t \,. \label{eq:stindepSDE}
\end{equation}
After replacing $V_t = \frac{\sin(\sqrt{2 \theta_t \alpha} Z_t) + 1}{2} - p_t$ in (\ref{eq:stindepSDE}), we obtain that the  process $Z$ satisfies the SDE
\begin{align}
dZ_t & = \left[ \frac{- \theta_t (\sin(\sqrt{2 \theta_t \alpha} Z_t) + 1 - 2 p_t)}{\sqrt{2 \theta_t \alpha} \cos{(\sqrt{2 \theta_t \alpha} Z_t)}} + \frac{\sqrt{2 \theta_t \alpha} \sin{(\sqrt{2 \theta_t \alpha} Z_t)}}{2 \cos{(\sqrt{2 \theta_t \alpha} Z_t)}} \right] \,dt + dW_t \nonumber \\
&  = \left[  \frac{\sqrt{2 \theta_t} \left( \sin(\sqrt{2 \theta_t \alpha} Z_t) (\alpha - 1) + (2 p_t -1 ) \right)}{2 \sqrt{\alpha} \cos{(\sqrt{2 \theta_t \alpha} Z_t)}}  \right] \,dt + dW_t \,. 
\end{align}

@@

We can see in Figure (\ref{fig:skew_after_lamperti}) the effect of the Lamperti transformation upon the forecast error data. 
The Lamperti transformation has greatly reduced the forecast error skewness, ensuring that the process stays in the range $[0,1]$. Therefore, in this case, the transition densities of the process Z can be adequately approximated through Gaussian densities.
%---END SECTION 3---

@@

Suppose that any of $M$ non-overlapping paths of the continuous-time It\^{o} process $V = \{ V_t, t >0 \}$ is sampled at $N$ equispaced discrete points with length interval $\Delta_N$, and let $ V^{M,N}=\{ V_{t_1^{M,N}} , V_{t_2^{M,N}} ,\ldots , V_{t_N^{M,N}} \}$ denote this random sample.

@@

We compare two candidate models to find the best-fit that maximizes the retained information,
\begin{itemize}
  \item Model 1: This model does not feature derivative tracking.
  \begin{align}
  dX_t &=  - \theta_0 (p_t-X_t) \  dt + \sqrt{2 \theta \alpha X_t (1-X_t)} \  dW_t   \nonumber \\ %\quad t > 0
  X_0 & = x_0 \,, \label{M0}
 \end{align}
  with $\theta_t= \theta_0 $.

@@

  \item Model 2: This model features derivative tracking, i.e. it is equivalent to (\ref{model:derivative_tracking_X}).
  \begin{align}
  dV_t &=  - \theta_t V_t \  dt + \sqrt{2 \theta_t \alpha (V_t +p_t ) (1-V_t-p_t)} \  dW_t   \nonumber \\ %\quad t > 0
  V_0 & = v_0 \,, \label{M2}
  \end{align}
  with $\theta_t$ given by (\ref{theta_t}).
\end{itemize}

>>>>

\begin{itemize}
\item $a(\cdot; p_t, \dot{p}_t, \bm{\theta}): [0,1] \to \mathbb{R} $  denotes a drift function,
\item $b(\cdot; p_t, \dot{p}_t, \bm{\theta}): [0,1] \to \mathbb{R}_+ $  a  diffusion function,
\item $\bm{\theta}$ is a vector of unknown parameters,
\item $(p_t)_{t \in [0,T]}$ is a time-dependent deterministic function $[0,1]$-valued and $ (\dot{p}_t)_{t \in [0,T]}$ is its time derivative,
\item $\{W_t, \, t \in [0,T] \}$ is a standard real-valued Wiener process.
\end{itemize}

@@

Let $\bm{\theta} = (\theta_0,\alpha)$, and choose a state-dependent diffusion term that avoids the process exiting from the range $[0,1]$ as follows:
  \begin{equation}
    b (X_t; \bm{\theta} )= \sqrt{2 \alpha \theta_0 X_t (1 - X_t)}
  \end{equation}
  where $\alpha >0$ is a unknown parameter that controls the path variability. This diffusion term belongs to the Pearson diffusion family and, in particular, it defines a Jacobi type diffusion. It is useful to recall that (\cite[440]{foso}) a Pearson diffusion is a stationary solution to a stochastic differential equation of the form
 \begin{equation}
    dX_t = - \theta (X_t - \mu) dt + \sqrt{2 \theta (a X_t^2 + b X_t + c)} dW_t
  \end{equation}
where $\theta>0$, and $a$, $b$ and $c$ are parameters such that the square root is well defined when $X_t$ is in the state space. These parameters, together with $\mu$, the mean of the invariant distribution, determine the state space of the diffusion as well as the shape of the invariant distribution.

An exhaustive classification of the (stationary) Pearson diffusions is presented in (\cite[440-443]{foso}) where, in particular, it is discussed the case $a < 0$ and $b(x; \bm{\theta}) = \sqrt{2 a \theta x (x-1)}$, where the invariant distribution is a Beta distribution with parameters $\left( \frac{\mu}{-a}, \frac{1 - \mu}{-a} \right),$ that leads to the well-known Jacobi diffusions, so-called because the eigenfunctions of the infinitesimal generator of these processes are the Jacobi polynomials (see, for example, \cite[2860-2861]{leph}).

It is worth mentioning that Jacobi diffusions have been successfully applied in several disciplines, among them finance (see \cite{vago} and references therein) and neuroscience (\cite{dotala}).

However, a distinctive feature in our proposed model 
\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dX_t \!\!\!&= (\dot{p}_t  - \theta_t (X_t - p_t) ) dt +\sqrt{2 \alpha \theta_0 X_t (1 - X_t)} dW_t\,, \quad t \in [0,T]  \\
   X_0  \!\!\!&=  x_0\in [0,1] \,,
 \end{array}\right.  \label{ourmodel}
\end{equation}

is that the drift term contains the time-varying parameter $\theta_t$, rendering the solution $X$ to (\ref{ourmodel}) a non-stationary and time-inhomogeneous process. To ensure that the process $X$ be the unique strong solution to (\ref{ourmodel}) for all $t \in [0,T]$ with state space $[0,1]$ a.s., the mean-reversion time-varying parameter must satisfy the following condition:
%has to be selected according to the following rule. Observe that the time derivative term $\dot{p}_t $ is not controlled to maintain that $X_t$ stays a.s. inside the range $[0,1]$. In other words, the zero drift line defined by $a(x; p_t,\bm{\theta}) =0$, which is an attractor, must be contained inside the range $[0,1]$. Thus, we must have that
%\begin{equation}
%\frac{- |\dot{p_t}|}{p_t} \leq \theta_t \leq \frac{|\dot{p}_t|}{1- p_t}
%\end{equation}
%which is satisfied  by choosing a time-dependent  $\theta_t$ as follows,
%\begin{equation}
%\theta_t = \max \left( \theta_0 \ , \ \frac{|\dot{p}_t|}{\min (p_t, 1-p_t)}  \right ),  \quad \theta_0 >0\,. \label{theta_t}
%\end{equation}
\begin{equation}\label{Assumption:3}
\theta_t\geq \max\left(\frac{\alpha\theta_0+\dot p_t}{1-p_t},\frac{\alpha\theta_0-\dot p_t}{p_t}\right)\tag{C}. 
\end{equation} \label{condC}
The proof of this theoretical statement is presented in Section \ref{Appendix}. \\

Remark: Condition (C) shows that the time-varying parameter $\theta_t$ becomes unbounded when $p_t = 0$ or $p_t = 1$. Therefore, we consider the following truncated prediction function
\begin{equation}
  p_t^{\epsilon} = \left\{
  \begin{array}{@{}cl@{}}
    \epsilon \!\!\!&\text{if } p_t  < \epsilon  \\
    p_t  \!\!\!&\text{if }  \epsilon \leq p_t < 1 - \epsilon \\
    1 - \epsilon  \!\!\!&\text{if }  p_t \geq 1 - \epsilon
 \end{array}\right.  \label{corrforecast}
\end{equation}
that satisfies $p_t^{\epsilon} \in [\epsilon, 1 - \epsilon]$ for any $0 < \epsilon < \frac{1}{2}$ and $t \in [0,T]$, providing that $\theta_t$ is bounded for every $t \in [0,T]$. \\
From now on, we will keep the notation $p_t$ to denote the truncated prediction function (\ref{corrforecast}) with, unless specified otherwise, $\epsilon = 10^{-2}$.

@@

%\subsubsection*{Change of Variables:}
\subsection{A model specification for the forecast error}
 After applying to (\ref{ourmodel}) the simple change of variables $$V_t = X_t - p_t \,,$$ we may introduce the following model for the forecast error of the normalized wind power production:

\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dV_t \!\!\!&=  - \theta_t V_t  \, dt + \sqrt{2 \alpha \theta_0 (V_t +p_t ) (1-V_t-p_t)} \, dW_t\,, \quad t \in [0,T]  \\
   V_0  \!\!\!&=  v_0\in [-1 + \delta,1 - \delta] \,.
 \end{array}\right.  \label{VtSDE}
\end{equation}

@@

We consider the following Lamperti transformation  
%{\color{red} (Remind that continuous-time models with different diffusion term are singular and the likelihood cannot be written down. Besides, it can be unstable to work with low frequency. If the frequency of the data is too low, the Lamperti transform helps to detect this problem (in principle, it is not clear how to determine the right frequency for estimation. )}
\begin{equation}
Z_t = h(V_t, t) = \frac{1}{\sqrt{2 \theta_0 \alpha}} \int_{\xi}^{x} \frac{1}{\sqrt{(u + p_t)(1 - u - p_t)}} \,du \Bigg{\vert}_{x = V_t} \,,
\end{equation}
where $\xi$ is an arbitrary point of the state space of the process $V$. The choice of $\xi = \frac{1}{2} - p_t$, where $p_t$ is a known input, leads to the process 
\begin{equation}
Z_t = h(V_t, t) = \frac{1}{\sqrt{2 \theta_0 \alpha}} \arcsin (2(V_t + p_t) -1)\,,
\end{equation}
that, after applying It\^{o}'s formula on $h(V_t, t)$, gives the state-independent diffusion SDE
\begin{equation}
dZ_t = \left[  \frac{- \theta_t V_t}{ \sqrt{2 \theta_t \alpha (V_t + p_t)(1 - V_t - p_t)}} - \frac{1}{4} \frac{\sqrt{2 \theta_t \alpha} \left( 1 - 2 (V_t + p_t)\right)}{\sqrt{(V_t + p_t)(1 - V_t - p_t)}}  \right] \,dt + dW_t \,. \label{eq:stindepSDE}
\end{equation}
After replacing $V_t = \frac{\sin(\sqrt{2 \theta_t \alpha} Z_t) + 1}{2} - p_t$ in (\ref{eq:stindepSDE}), we obtain that the  process $Z$ satisfies the SDE
\begin{align}
dZ_t & = \left[ \frac{- \theta_t (\sin(\sqrt{2 \theta_t \alpha} Z_t) + 1 - 2 p_t)}{\sqrt{2 \theta_t \alpha} \cos{(\sqrt{2 \theta_t \alpha} Z_t)}} + \frac{\sqrt{2 \theta_t \alpha} \sin{(\sqrt{2 \theta_t \alpha} Z_t)}}{2 \cos{(\sqrt{2 \theta_t \alpha} Z_t)}} \right] \,dt + dW_t \nonumber \\
&  = \left[  \frac{\sqrt{2 \theta_t} \left( \sin(\sqrt{2 \theta_t \alpha} Z_t) (\alpha - 1) + (2 p_t -1 ) \right)}{2 \sqrt{\alpha} \cos{(\sqrt{2 \theta_t \alpha} Z_t)}}  \right] \,dt + dW_t \,. 
\end{align}

@@

{\color{red}We can see in Figure (\ref{fig:skew_after_lamperti}) the effect of the Lamperti transformation upon the forecast error data. 
The Lamperti transformation has greatly reduced the forecast error skewness, ensuring that the process stays in the range $[0,1]$. Therefore, in this case, the transition densities of the process Z can be adequately approximated through Gaussian densities.}
%---END SECTION 3---

@@

Suppose that any of $M$ non-overlapping paths of the continuous-time It\^{o} process $V = \{ V_t, t  \in [0,T] \}$ is sampled at $N$ equispaced discrete points with length interval $\Delta_N$, and let $ V^{M,N}=\{ V_{t_1^{M,N}} , V_{t_2^{M,N}} ,\ldots , V_{t_N^{M,N}} \}$ denote this random sample.

@@

We compare two candidate models to find the best-fit that maximizes the retained information,
\begin{itemize}
  \item Model 1: This model does not feature derivative tracking.
\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dX_t \!\!\!&=  - \theta_0 (X_t - p_t) dt +\sqrt{2 \alpha \theta_0 X_t (1 - X_t)} dW_t\,, \quad t \in [0,T]  \\
   X_0  \!\!\!&=  x_0\in [0,1] \,,
 \end{array}\right.  \label{Model1}
\end{equation}
 with $\theta_0 > 0, \, \alpha > 0$.

@@

  \item Model 2: This model features derivative tracking and time-varying mean-reversion parameter:  
%i.e. it is equivalent to (\ref{model:derivative_tracking_X}).
%  \begin{align}
%  dV_t &=  - \theta_t V_t \  dt + \sqrt{2 \theta_t \alpha (V_t +p_t ) (1-V_t-p_t)} \  dW_t   \nonumber \\ %\quad t > 0
%  V_0 & = v_0 \,, \label{M2}
%  \end{align}
%  with $\theta_t$ given by (\ref{theta_t}).
\begin{equation}
  \left\{
  \begin{array}{@{}rl@{}}
    dX_t \!\!\!&= (\dot{p}_t  - \theta_t (X_t - p_t) ) dt +\sqrt{2 \alpha \theta_0 X_t (1 - X_t)} dW_t\,, \quad t \in [0,T]  \\
   X_0  \!\!\!&=  x_0\in [0,1] \,,
 \end{array}\right.  \label{Model2}
\end{equation}
 with $\theta_0 > 0, \, \alpha > 0$ and .
\end{itemize}
